% !TeX root = master.tex

\section{Introduction}

RaceTools is a small collection of race-day tools for sailors. The design goal is
deliberately narrow: show only the information you actually need, with stable numbers and
clear UI that remains readable when the start line is busy and you are operating the
phone one-handed.

The app opens on a Home screen with three tools: RaceStarter, RacePerformance, and RaceLifter.
Boat and Settings are accessed from Home so all modes share the same configuration.

This document is a technical report that explains how the app works and why the math is
set up the way it is. The report is structured by tool. The RaceStarter part focuses on
start line setup and the race view, and on the estimation methods needed to make distance,
speed, and direction stable enough to trust. The RacePerformance part explains the VMG evaluator:
it builds a low-pass baseline for speed and heading, compares each new sample against that
baseline using an assumed TWA for beat or run (or zero for reach), and shows the resulting
percent change as a time history plot. The RaceLifter part explains the heading history
plot and how its bins, mean, and scaling are computed.

Across both tools, the most important requirement is trust. Raw GPS fixes are jittery, and
naively differentiating them produces even noisier speed and heading. For this reason, we
invest most of the mathematical complexity in estimation: turning intermittent, noisy
measurements into a smooth position and a stable velocity vector. When that layer works
well, everything above it becomes straightforward: geometry on top of a steady state
estimate, rather than UI that fights noisy sensors.
